Longitudinal Language Analysis in the Diagnosis of Mild Cognitive Impairment: A Case Study

### 2019-03-05 Source/Presidents Corpus Paper/paper1.tex
changes to paper1

\author{
  Jomar C. Alcantara \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences\\
  Aston University\\
  Birmingham, UK \\
  \texttt{alcantaj@aston.ac.uk} \\
  \And
  Peter Sawyer \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences \\
  Aston University\\
  Birmingham, UK \\
  \texttt{p.sawyer@aston.ac.uk} \\
  \And
  Cristina Romani \\
  Department of Psychology \\
  School of Life and Health Science\\
  Aston University\\
  Birmingham, UK \\
  \texttt{c.romani@aston.ac.uk} \\
}
% HEADER ENDS
% DOCUMENT BEGINS
\begin{document}
\maketitle
\begin{abstract}
Dementia is a neurological disorder that has been consistently growing in numbers over time. The analysis of language has been highlighted as a potential method for diagnosing Alzheimer's disease at its early or pre-clinical stages. This paper builds on the work of Berisha and Liss who looked at extracting language features from the non-scripted portions of press conferences from two US presidents, President Ronald Reagan, who would later go on to receive a diagnosis of Alzheimer's Disease and President George Herbert Walker Bush who received no such diagnosis. In this paper, we add the transcripts of President Donald John Trump who in terms of age is a much more appropriate match and we extract a larger variety of language features which have been shown to be associated with language decline due to Dementia. Once the features were extracted we completed a regression analyses. We found that President Reagan showed a significant decrease in unique words over time and his usage of nouns and adjectives. We also found that there was an increase in his usage of non-specific nouns, adverbs and pronouns. We found no significant trends in the transcripts of Presidents Bush and Trump. This study shows that language deterioration can be tracked over time in those with Mild Cognitive Impairment and is generally differentiable from healthy controls. The results also potentially demonstrate the utility of longitudinal language analysis as a method of diganosing Mild Cognitive Impairment in the future.
\end{abstract}
\keywords{Mild Cognitive Impairment \and Alzheimer's Disease \and Natural Language Processing \and Longitudinal \and Early Diagnosis \and Language}

\section{Background}

@@

I took 46 transcripts of Ronald Reagan’s (RR) press conferences from 1981 to 1988 and compared them with 134 press conferences by George H. W. Bush (GHWB) and 29 press conferences conducted by Donald J. Trump (DJT).  I analyzed transcripts for lexical features shown to change longitudinally with dementia  (for a comprehensive review of these, see the literature review above). For this collection of documents, I generated a number of features which looked at a number of different aspects of each document. These features encompassed, word level, sentence level and document level features and included a number of features contained in the study by Berisha and Liss with the aim of replicating and extending on their findings. These findings were. number of unique words, non-specific nouns and fillers and low imageability (LI) verbs. Imageability is characterized, according to Berisha and Liss, as the ease with which a term gives rise to a sensory .mental image. I compared the trends described in the transcripts of RR and GHWB, but also included DJT. Berisha and Liss originally made the comparison as it GHWB (GHWB - age at the start of presidency - 64 years, 222 days) was the closest match to RR in terms of age (RR  - age at the start of presidency, 69 yeaars and 349 days). However, with the inauguration of Trump, he now is the closest comparible president in terms of age (DJT - age at the start of presidency - 70 years, 220 days). It would be interesting to look at a comparison of RR and DJT to see whether the comparisons made by Berisha and Liss hold true with this more appropriate match (in terms of age). DJT as with GHWB has no known diagnosis of AD. I used the press conference transcripts in the American Presidency Project (APP) archive as a data source for this project. The APP is a comprehensive and organized searchable database of presidential documents, including transcripts of speeches, transcripts of news conferences, and other public documents.

@@

To generate the files necessary for analysis, I downloaded each transcript and performed the following changes. I omitted the prepared statement by the president and any speech by other individuals. I started each transcript at the beginning of the first answer to a question. I filtered any annotations that were added to the transcript, including any references or clarifications, and any laughter. It's worth noting that there appears to be a difference in how 'hesitations' were marked down between each president, for RR hesitations were marked by a single hyphen whereas for GHWB hesitations are marked by a double hyphen. In order to maintain consistency when parsing through the documents, I have changed both types of hesitation to be marked by a single hyphen. I also omitted one word sentences as this data would, from a theoretical perspective, not be relevant for language analysis. I did not control for the length of the document, but generated features which would normalise by the length of the document. I therefore was able to include all press conferences by both Ronald Reagan and George H.W. Bush where there was a question and answer session conducted at least in part by the sitting president (2 press conferences of GHWB were ommited due to a lack of a question and answer session).

>>>>

\author{
  Jomar C. Alcantara \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences\\
  Aston University\\
  Birmingham, UK \\
  \texttt{alcantaj@aston.ac.uk} \\
  \And
  Peter Sawyer \\
  Department of Computer Science \\
  School of Engineering and Applied Sciences \\
  Aston University\\
  Birmingham, UK \\
  \texttt{p.sawyer@aston.ac.uk} \\
  \And
  Cristina Romani \\
  Department of Psychology \\
  School of Life and Health Science\\
  Aston University\\
  Birmingham, UK \\
  \texttt{c.romani@aston.ac.uk} \\
}
% HEADER ENDS
% DOCUMENT BEGINS
\begin{document}
\maketitle
\begin{abstract}
Dementia is a neurological disorder that has been consistently growing in numbers over time. The analysis of language has been highlighted as a potential method for diagnosing Alzheimer's disease at its early or pre-clinical stages. This paper builds on the work of Berisha and Liss who looked at extracting language features from the non-scripted portions of press conferences from two US presidents, President Ronald Reagan, who would later go on to receive a diagnosis of Alzheimer's Disease and President George Herbert Walker Bush who received no such diagnosis. In this paper, we add the transcripts of President Donald John Trump who in terms of age is a much more appropriate match and we extract a larger variety of language features which have been shown to be associated with language decline due to Dementia. Once the features were extracted we completed a regression analyses. We found that President Reagan showed a significant decrease in unique words over time and his usage of nouns and adjectives. We also found that there was an increase in his usage of non-specific nouns, adverbs and pronouns. We found no significant trends in the transcripts of Presidents Bush and Trump. This study shows that language deterioration can be tracked over time in those with Mild Cognitive Impairment and is generally differentiable from healthy controls. The results also potentially demonstrate the utility of longitudinal language analysis as a method of diganosing Mild Cognitive Impairment in the future.
\end{abstract}
\keywords{Cognitive Dysfunction \and Alzheimer Disease \and Natural Language Processing \and Early Diagnosis \and Speech}

\section{Introduction}

@@

We took 46 transcripts of Ronald Reagan’s (RR) press conferences from 1981 to 1988 and compared them with 134 press conferences by George H. W. Bush (GHWB) and 29 press conferences conducted by Donald J. Trump (DJT).  We analyzed transcripts for lexical features shown to change longitudinally with dementia. For this collection of documents, We generated a number of features which looked at a number of different aspects of each document. These features encompassed, word level, sentence level and document level features and included a number of features contained in the study by Berisha and Liss with the aim of replicating and extending on their findings. These features were: number of unique words, non-specific nouns and fillers and low imageability (LI) verbs. Imageability is characterized, according to Berisha and Liss, as the ease with which a term gives rise to a sensory mental image. We compared the trends described in the transcripts of RR and GHWB, but also included DJT. Berisha and Liss originally made the comparison as it GHWB (GHWB - age at the start of presidency - 64 years, 222 days) was the closest match to RR in terms of age (RR  - age at the start of presidency, 69 yeaars and 349 days). With the inauguration of Trump, he now is the closest comparible president in terms of age (DJT - age at the start of presidency - 70 years, 220 days). It would be interesting to look at a comparison of RR and DJT to see whether the comparisons made by Berisha and Liss hold true with this more appropriate match (in terms of age). DJT as with GHWB has no known diagnosis of AD. We used the press conference transcripts in the American Presidency Project (APP) archive as a data source for this project. The APP is a comprehensive and organized searchable database of presidential documents, including transcripts of speeches, transcripts of news conferences, and other public documents.

@@

To generate the files necessary for analysis, We downloaded each transcript and performed the following changes. We omitted the prepared statement by the president and any speech by other individuals. We started each transcript at the beginning of the first answer to a question. We omitted any annotations that were added to the transcript, including any references or clarifications, and any laughter. It's worth noting that there appears to be a difference in how 'hesitations' were marked down between each president, for RR hesitations were marked by a single hyphen whereas for GHWB and DJT hesitations are marked by a double hyphen. In order to maintain consistency when parsing through the documents, we changed both types of hesitation to be marked by a single hyphen. We also omitted one word sentences as this data would, from a theoretical perspective, not be relevant for language analysis. We did not control for the length of the document, but generated features which would subsequently be normalised by the length of the document. We therefore were able to include all press conferences by both Ronald Reagan and George H.W. Bush where there was a question and answer session conducted at least in part by the sitting president (2 press conferences of GHWB were ommited due to a lack of a question and answer session).
